---
title: "Data Management -- not to be submitted"
output:
  pdf_document:
    keep_tex: true
geometry: margin=1.5cm
header-includes:
   - \usepackage{booktabs}
   - \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# You don't have to submit this lab

# Introduction

I have downloaded three data sets from gapminder.org, located in the data folder (they have informative titles):

 * co2_emissions_tonnes_per_person.csv
 * energy_production_per_person.csv
 * energy_use_per_person.csv

```{r, message = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(purrr)
library(GGally)
```

# Read in the data sets and take a first look at them

You'll end up with three separate data frames.  Use the `read_csv` function from the readr package and the `View` function.

```{r}
co2 <- read_csv("data/co2_emissions_tonnes_per_person.csv")
View(co2)
energy_production <- read_csv("data/energy_production_per_person.csv")
View(energy_production)
energy_use <- read_csv("data/energy_use_per_person.csv")
View(energy_use)
```

# Pivot the data frames so that they have columns for country and year, and a column for co2, energy_production, and energy_use respectively.

Use the `pivot_longer` function from the tidyr package.

```{r}
co2 <- co2 %>%
   pivot_longer(-country, names_to = "year", values_to = "co2")
energy_production <- energy_production %>%
   pivot_longer(-country, names_to = "year", values_to = "energy_production")
energy_use <- energy_use %>%
   pivot_longer(-country, names_to = "year", values_to = "energy_use")
```

# Which 10 countries emitted the most co2 in 2010?  Which 10 countries produced the most energy in 2010?  Which 10 countries used the moest energy in 2010?

Use the `filter` function from the dplyr package to select just the data from 2010, and then use either the `arrange` function and the `tail` function, or the `top_n` function (both from the dplyr package) to view the top 10.  Recall that you can use the `drop_na` function to drop rows with missing values.

```{r}
co2 %>% filter(year == 2010) %>% drop_na() %>% arrange(co2) %>% tail(10)
co2 %>% filter(year == 2010) %>% top_n(10, co2)
energy_production %>% filter(year == 2010) %>% drop_na() %>% arrange(energy_production) %>% tail(10)
energy_production %>% filter(year == 2010) %>% top_n(10, energy_production)
energy_use %>% filter(year == 2010) %>% drop_na() %>% arrange(energy_use) %>% tail(10)
energy_use %>% filter(year == 2010) %>% top_n(10, energy_use)
```

# Which 10 countries emitted the least co2 in 2010?

How could you do this using `top_n` by modifying the co2 variable so that that the "top n" values correspond to the countries with lowest emissions?

```{r}
co2 %>% filter(year == 2010) %>% drop_na() %>% arrange(co2) %>% head(10)
co2 %>% filter(year == 2010) %>% top_n(10, -1 * co2)
```

# Merge the three data sets into one data set.  View the combined data to make sure it worked.

This is necessary to examine relationships between co2 and energy use and production

You will want to use one of the join family of functions (see `?join`).  You will need two calls to a join function, the first to join two of the data sets, and the second call to add the third data set in.

```{r}
all_data <- co2 %>%
   full_join(energy_production) %>%
   full_join(energy_use)
View(all_data)
```

# Create a plot of co2 emissions for 2010 vs energy production for 2010, and a second plot showing co2 emissions for 2010 vs energy use for 2010.  Which relationship is stronger?

```{r}
ggplot(data = all_data %>% filter(year == 2010),
       mapping = aes(x = energy_production, y = co2)) +
   geom_point()

ggplot(data = all_data %>% filter(year == 2010),
       mapping = aes(x = energy_use, y = co2)) +
   geom_point()
```
